### 2020-01-24 paper/sex_selection_spacing-ver3.tex
Separate out introduction of sex-selection
More concise version of introduction and more upfront discussion of
birth spacing.

India has changed significantly over the last three decades:
the economy has grown substantially,
education has increased for both males and females,
and the total fertility rate has fallen to the point where, at 2.2, it is just
above replacement 
\citep{Bosworth2008,Guilmoto2013,Dharmalingam2014,International-Institute-for-Population-Sciences-IIPS2017}.
However, access to prenatal sex determination also expanded rapidly starting in the mid-1980s,
and, combined with a continued strong son preference, this dramatically increased the 
males-to-females ratio at birth 
\citep{das_gupta97,Sudha1999,Arnold2002,retherford03b,jha06,Guilmoto2009a,Guilmoto2012,
Bongaarts2013,Portner2015b,Jayachandran2017}.

What has not previously been examined is the impact of these changes on birth spacing.
Economic growth, increasing female education, and falling fertility has been
shown to significantly affect birth spacing in other countries [TK references].
India is, however, fundamentally different in at least one respect and that is 
the substantial use of sex-selective abortions.

The stronger the son preference and the lower the desired fertility, the more likely 
couples are to resort to sex-selective abortions, and each abortion increases the 
interval between births by six to twelve months.%
\footnote{
The increase consists of three parts.
First, starting from the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Hence, the changes in birth spacing may be much larger in India than what we have
observed in other places as they developed.

Understanding how birth spacing changes is important for three reasons.

First, tempo effects, arising from changes in birth spacing, biases the total fertility 
rate (TFR), in the sense that the TFR becomes a more biased measure of cohort fertility
\citep{Bongaarts1999,Ni-Bhrolchain2011}.
Hence, it is possible that fertility is, in fact, substantially higher than what
is generally accepted based on the TFR.

@@

Finally, birth spacing has long served as a measure of son preference, with strong
son preference associated with shorter birth intervals the fewer sons a family has 
\citep{ben-porath76b,Leung1988}.
Instead, longer birth intervals in the absence of sons, which previously would be 
taken as an indicator of lower son preference, may now instead arise because couples 
with a strong son preference use sex selection.

Here, I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
I apply an empirical method that directly incorporates the effects of 
sex-selective abortions on birth intervals
\emph{and} 
the likelihood of a son. 
The method can be used to analyze both situations with and without prenatal
sex selection.
I apply the method to birth histories of Hindu women, using data from the four
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2016. 
I address three main questions.

% [Hypotheses / questions]

First, how did average birth intervals change over time for different groups and how do 
these changes relate to the use of sex selection?
The hypothesis is that in the absence of sex selection son preference leads to shorter 
spacing the fewer boys a family has, whereas son preference increases spacing the fewer 
boys a family has when prenatal sex selection is available, possibly to the point where
we observe a reversal of the standard spacing pattern.
Families with son preference who---for one reason or another---do not use prenatal sex 
selection likely continue to have shorter birth spacing.
Based on the prior literature on sex selection, I focus on how birth intervals have
changed for different education levels, urban or rural, and the sex composition of 
prior children.

Second, have the shapes of the survival curves changed?
Although examining average birth intervals helps us understand the overall changes in 
spacing behavior, it tells us little about how these changes came about.
For example, the motivations that lead to an increase in average spacing are different if 
the increase comes mainly from parents trying to avoid very short birth intervals or from 
continued use of sex selection to ensure a son.

>>>>

India has changed significantly over the last three decades:
the economy has grown substantially,
education levels have increased for both males and females,
and the total fertility rate has fallen to the point where, at 2.2, it is just
above replacement 
\citep{Bosworth2008,Guilmoto2013,Dharmalingam2014,
International-Institute-for-Population-Sciences-IIPS2017}.

Over the same period, India has, however, also experienced a rapidly expanding 
access to prenatal sex determination.
Combined with continued strong son preference, the result has been a dramatic 
increase in the males-to-females ratio at birth
\citep{das_gupta97,Sudha1999,Arnold2002,retherford03b,jha06,Guilmoto2009a,Guilmoto2012,
Bongaarts2013,Portner2015b,Jayachandran2017}.

What the prior research on India has not appreciated is that since each abortion 
increases the interval between births by six to twelve months, the growing use of 
sex-selective abortions has the potential to increase birth spacing substantially. %
\footnote{
The increase consists of three parts.
First, starting from the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Add to this increase that the combination of increasing female education, 
lower fertility, higher household income, and the low and declining female 
labor force participation are also likely to impact birth spacing.
Hence, the changes in birth spacing may be much more substantial in India 
than what we have observed in other countries as they developed.

Understanding the changes in birth spacing is essential for three reasons.

First, tempo effects, arising from changes in birth spacing, bias the total 
fertility rate (TFR), in the sense that the TFR becomes a more biased measure 
of cohort fertility \citep{Bongaarts1999,Ni-Bhrolchain2011}.
Hence, fertility may be, in fact, substantially higher than what is generally 
accepted based on the TFR.

@@

Finally, birth spacing has long served as a measure of son preference, with strong
son preference associated with shorter birth intervals, the fewer sons a family has 
\citep{ben-porath76b,Leung1988}.
Now, longer birth intervals in the absence of sons, rather than an indicator of 
lower son preference, may instead arise because couples with a strong preference 
for sons use sex selection.

Here, I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
I apply a competing risk hazard model that directly incorporates the effects of 
sex-selective abortions on birth intervals
\emph{and} 
the likelihood of a son, 
but which also works in cases without prenatal sex selection.
I apply the method to birth histories of Hindu women, using data from the four
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2016.

% [Hypotheses / questions]
I address four main questions.
[TK combine questions and highlight of answers here?]

First, how did average birth intervals change over time for different groups and how do 
these changes relate to the use of sex selection?
The hypothesis is that in the absence of sex selection son preference leads to shorter 
spacing the fewer boys a family has, whereas son preference increases spacing the fewer 
boys a family has when prenatal sex selection is available, possibly to the point where
we observe a reversal of the traditional spacing pattern.
Families with son preference who---for one reason or another---do not use prenatal sex 
selection likely continue to have shorter birth spacing.
Based on the prior literature on sex selection, I focus on how birth intervals have
changed for different education levels, urban or rural, and the sex composition of 
prior children.

Second, how has the distribution of birth spacing changed?
Although examining average birth intervals helps us understand the overall changes in 
spacing behavior, it tells us little about how these changes came about.
For example, the motivations that lead to an increase in average spacing are different if 
the increase comes mainly from parents trying to avoid very short birth intervals or from 
continued use of sex selection to ensure a son.

Third, to what extent does TFR reflect cohort fertility?
An advantage of the empirical model is that I can predict fertility behavior taking into
account both the likelihoods of parity progression and having a child of a specific sex.
I compare those predictions with fertility rate estimates based on the same data to
show how the increases in birth spacing has led to a substantial overestimate of the
decline in fertility [TK too imprecise? underestimate of the level, rather?]

Finally, has the relationship between birth spacing and mortality changed over time?
[TK]
